---
title: "Bioconductor: R for omics data"
subtitle: "Week 13 -- lecture B"
pagetitle: "w13a: Bioconductor"
date: 2025-11-15
author:
  - name: Menuka Bhandari
    #orcid: 0000-0002-3514-7462
    email: bhandari.72@osu.edu
    affiliation: 
      - name: CFAES Bioinformatics Core, Ohio State University
        url: https://mcic.osu.edu/bioinformatics
  - name: Jelmer Poelstra
    orcid: 0000-0002-3514-7462
    email: poelstra.1@osu.edu
editor_options: 
  chunk_output_type: console
other-links:
  - text: "Week 13 overview"
    icon: book
    href: w13_overview.html
  - text: "w13b: Bioconductor: R for omics data"
    icon: bookmarks
    href: w13b_bioconductor.html
execute:
  eval: false
---

--------------------------------------------------------------------------------

<br>

## Introduction
R packages are available in many places. In our previous lecture, we learnt to install R packages through CRAN using `install.packages()` function.
In this lecture, we will learn about [Bioconductor](https://bioconductor.org/), a specialized repository
for **bioinformatics packages in R** that focuses on analysis of high-throughput genomic data. 
initially developed for the analysis of microarray data.  Bioconductor packages 
are used to analyze a wide range of omics data types such as bulk and single-cell RNA-seq, Chip seq, copy number analysis, micro array methylation and 
classic expression analysis,  flow cytometry, and many other domains. 
One of the key requirement of Bioconductor packages is to have high level documentation,
and workflows. The packages available in Bioconducter are provided [here](https://bioconductor.org/packages/release/BiocViews.html#___Software).


### Overview & learning goals

In this lecture, you will learn : 

- Basics of Bioconductor
- How to install and load Bioconductor packages
- Use DECIPHER package to align the genomic sequences

Specifically, you will learn:

- Install package using `BiocManager::install()` function
- Align the genomic seqeunces

## Set up and Data preparation

### Open a new Quarto document in Rstudio

Open a new Quarto file `File` \> `New File` \> `Quarto Document` and save it inside the `week13` directory as 
`week13b_bioconductor`.

## Packages available in Biocondctor

There are 4 different types of software packages available in Bioconductor:

1. Software packages for the analysis and comprehension of high-throughput genomic data.

2. Annotation data packages that provide mappings between different biological identifiers and genomic coordinates.

3. Experiment data packages that contain example data sets for testing and demonstrating analysis methods.

4. Workflows and vignettes that provide step-by-step guides for common bioinformatics analyses.

BioConductor delivers releases where a set of packages is published at once and intended for 
compatibility only with a certain version of R. This is in contrast to CRAN where 
packages are added continuously with no reference to particular versions of R.
Bioconducter page provide guidance on package acquisition,
create the documents that are computable and better software engineering.

Bioconductor Help

1. Bioconductor Packages

2. biocViews

3. Bioconductor Forums




## Install Bioconductor packages

Similar to CRAN, Bioconductor packages needs to be installed once.  However, packages are installed slightly differently in Bioconductor.
To install Bioconductor packages, you first need to install the `BiocManager` package from
CRAN (if you haven't already done so). You can then use `BiocManager::install()` to install
specific Bioconductor packages.

### Install BiocManager 

We need to install BiocManager through which we can install Bioconductor packages. The command to install BiocManager is provided below:
```{r eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

```


### Install a Bioconductor package
Once the BiocManager is installed, use the following command to install few packages of Bioconductor
e.g., GenomicRanges, DECIPHER

```{r eval=FALSE}
## Install packages
BiocManager::install("GenomicRanges")
BiocManager::install("DECIPHER")
```

### Load the packages
You should be an expert now in loading the packages in R since we have loaded so many packages since `week11`.
Use the `library()` function to load the Bioconductor package in a similar way as CRAN packages.
```{r}
### Load the DECIPHER package
library(DECIPHER)
```


##BiocManager::install(update = TRUE, ask = FALSE)

::: {.callout-tip}
## Package movement from CRAN and Bioconductor

Packages can become available on CRAN after being available in Bioconductor and vice versa.
what is the reason for it
`vcfR` package was first available in 
:::

This lecture, we will introduce you to Bioconductor, a powerful 
open-source software project that provides tools for the analysis and 
comprehension of high-throughput genomic data.

Bioconductor is a source project and repository for bioinformatics packages
in the R programming language. Since its start in 2001, Bioconductor package has grown 
to include more than 2000 packages contributed by a diverse community of researchers worldwide. 
The latest release of Bioconductor (version 3.22, released in October 2025) is compatible with R version 4.5.0.

## What type of packages are available in Bioconducter?
Bioconductor hosts a wide variety of packages that cater to different aspects of bioinformatics and computational biology. 

Some of the main types of packages available in Bioconductor include:

1. Software packages for the analysis and comprehension of high-throughput genomic data.
2. Annotation data packages that provide mappings between different biological identifiers and genomic coordinates.
3. Experiment data packages that contain example datasets for testing and demonstrating analysis methods. 
4. Workflows and vignettes that provide step-by-step guides for common bioinformatics analyses.



BiocViews: task view for package discovery
BiocViews is a controlled vocabulary for describing the content and functionality of Bioconductor packages.
### Overview & learning goals

## Example packages
- Biostrings : Read, write, search, and manipulate DNA, RNA, or amino acid sequences
- DECIPHER : 
 
- Sequence databases: import, maintain, view, and export a massive number of sequences.

- Sequence alignment: accurately align thousands of DNA, RNA, or amino acid sequences. Quickly find and
align the syntenic regions of multiple genomes.

- Oligo design: test oligos in silico, or create new primer and probe sequences optimized for a variety of objectives.

- Manipulate sequences: trim low quality regions, correct frameshifts, reorient nucleotides, determine 
consensus, or digest with restriction enzymes.

- Analyze sequences: find chimeras, classify into a taxonomy of organisms or functions, detect repeats, predict secondary structure, create phylogenetic trees, and reconstruct ancestral states.

- Gene finding: predict coding and non-coding genes in a genome, extract them from the genome, and export
them to a file.


Here, we will use a `DECIPHER` package for the alignment of the sequences. `DECIPHER` can align thousand of seqeunces in a very short time and maintain the accuracy

which 
MSA can show several things: 
- Is the regions conserved?

As the number of genomes for the alignment increases, the computational time increases as well.
It is a challenge to improve the performance when the number of sequences are added. 

## Bioconductor objects can be complex


## We will work on the Bioconducter packager DESEQ2 in the next lecture
